---
title: "MERFISH And CosMx Workflow"
format: html
execute:
  eval: false
  warning: false
  message: false
---

## 简介

MERFISH 和 CosMx 都是高分辨率成像平台，分析重点与 Visium 不同：细胞分割质量、panel 覆盖度、FOV / slide 批次、细胞类型标签和邻域结构更重要。本文件给出 Seurat v5 `ReadVizgen()` / `ReadNanostring()` 入口，并统一后续 QC、聚类、marker、邻域导出。

## Setup

```{r}
library(Seurat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(qs2)
library(here)
source(here::here("singlecell_utiles.R"))
```

## Project Configuration

```{r}
cfg <- list(
  platform = "MERFISH",  # "MERFISH" or "CosMx"
  merfish_dir = "/path/to/vizgen/output",
  cosmx_dir = "/path/to/cosmx/output",
  assay = "RNA",
  fov = "fov",
  sample_col = "sample",
  condition_col = "group",
  fov_col = "fov",
  celltype_col = "cell_type",
  min_counts = 20,
  min_features = 10,
  resolution = 0.4,
  dims = 1:30,
  marker_panel = c("EPCAM", "KRT8", "PTPRC", "COL1A1", "PECAM1"),
  results_dir = "Results/merfish_cosmx"
)

dir.create(cfg$results_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load Data

### Vizgen MERFISH

```{r}
if (cfg$platform == "MERFISH") {
  obj <- ReadVizgen(data.dir = cfg$merfish_dir, assay = cfg$assay)
  obj$sample <- basename(cfg$merfish_dir)
}
```

### Nanostring CosMx

```{r}
if (cfg$platform == "CosMx") {
  obj <- ReadNanostring(data.dir = cfg$cosmx_dir, assay = cfg$assay)
  obj$sample <- basename(cfg$cosmx_dir)
}

if (!cfg$condition_col %in% colnames(obj@meta.data)) obj[[cfg$condition_col]] <- "group_1"
if (!cfg$fov_col %in% colnames(obj@meta.data)) obj[[cfg$fov_col]] <- "fov_1"

obj
```

## QC

```{r}
count_col <- paste0("nCount_", cfg$assay)
feature_col <- paste0("nFeature_", cfg$assay)

qc_tbl <- obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  summarise(
    n_cells = n(),
    median_counts = median(.data[[count_col]], na.rm = TRUE),
    median_features = median(.data[[feature_col]], na.rm = TRUE),
    p95_counts = quantile(.data[[count_col]], 0.95, na.rm = TRUE)
  )
qc_tbl

VlnPlot(obj, features = c(count_col, feature_col), pt.size = 0, ncol = 2)
```

## QC And Segmentation Decision Table

```{r}
qc_decision_table <- tibble::tibble(
  metric = c(count_col, feature_col, "cell_boundary", "background_signal", "FOV_batch"),
  current_rule = c(
    paste(">=", cfg$min_counts),
    paste(">=", cfg$min_features),
    "manual segmentation review",
    "panel/platform-specific review",
    "composition and QC by FOV"
  ),
  reviewer_note = NA_character_
)

qc_decision_table
```

```{r}
obj <- subset(
  obj,
  cells = rownames(obj@meta.data)[
    obj@meta.data[[count_col]] >= cfg$min_counts &
      obj@meta.data[[feature_col]] >= cfg$min_features
  ]
)
```

## Normalization And Clustering

```{r}
DefaultAssay(obj) <- cfg$assay
obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj)
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = max(cfg$dims), verbose = FALSE)
obj <- FindNeighbors(obj, dims = cfg$dims)
obj <- FindClusters(obj, resolution = cfg$resolution)
obj <- RunUMAP(obj, dims = cfg$dims)

DimPlot(obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```

## Spatial Visualization

不同版本 Seurat 成像对象的 FOV 命名可能不同，先检查 `Images(obj)`。

```{r}
Images(obj)
ImageDimPlot(obj, fov = cfg$fov, group.by = "seurat_clusters", axes = TRUE)

panel_genes <- intersect(cfg$marker_panel, rownames(obj))
ImageFeaturePlot(obj, fov = cfg$fov, features = panel_genes, max.cutoff = "q95")
```

## Marker And Annotation

```{r}
Idents(obj) <- "seurat_clusters"
marker_tbl <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.1
)

marker_top <- marker_tbl |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 10, with_ties = FALSE) |>
  ungroup()
marker_top

annotation_map <- tibble::tibble(
  seurat_clusters = sort(unique(as.character(obj$seurat_clusters))),
  cell_type = paste0("CellType_", seurat_clusters),
  marker_support = NA_character_,
  reviewer_note = NA_character_
)

obj@meta.data <- obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  left_join(annotation_map, by = "seurat_clusters") |>
  tibble::column_to_rownames("cell")
```

## Annotation Decision Table

```{r}
annotation_decision_table <- annotation_map |>
  mutate(
    positive_markers = NA_character_,
    negative_markers = NA_character_,
    panel_limitation = NA_character_,
    spatial_location = NA_character_,
    confidence = NA_character_,
    use_for_downstream = NA
  )

annotation_decision_table
```

## FOV / Sample Composition

```{r}
composition_cols <- intersect(c(cfg$sample_col, cfg$celltype_col), colnames(obj@meta.data))
composition_tbl <- obj@meta.data |>
  count(across(all_of(composition_cols)), name = "n_cells") |>
  group_by(across(all_of(setdiff(composition_cols, cfg$celltype_col)))) |>
  mutate(frac_cells = n_cells / sum(n_cells)) |>
  ungroup()
composition_tbl
```

```{r}
fov_composition_tbl <- obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  count(.data[[cfg$sample_col]], .data[[cfg$condition_col]], .data[[cfg$fov_col]], .data[[cfg$celltype_col]], name = "n_cells") |>
  group_by(.data[[cfg$sample_col]], .data[[cfg$fov_col]]) |>
  mutate(frac_cells = n_cells / sum(n_cells)) |>
  ungroup()

fov_composition_tbl
```

## Neighborhood Hook

```{r}
neighbor_long <- NeighborComposition(obj, group_col = cfg$celltype_col, k = 20)
neighbor_summary <- neighbor_long |>
  group_by(group, neighbor_group) |>
  summarise(mean_neighbor_fraction = mean(fraction), .groups = "drop") |>
  arrange(desc(mean_neighbor_fraction))
neighbor_summary
```

## Platform Differences

```{r}
platform_review <- tibble::tribble(
  ~item, ~MERFISH, ~CosMx, ~review_note,
  "input_loader", "ReadVizgen", "ReadNanostring", "Use native Seurat v5 loader",
  "panel_bias", "panel-specific", "panel-specific", "Do not treat missing genes as absent biology",
  "segmentation", "cell boundary dependent", "cell boundary dependent", "Review failed / edge cells",
  "deconvolution", "skip", "skip", "Single-cell resolution; use neighborhood analysis instead"
)
platform_review
```

## Export

```{r}
readr::write_csv(qc_tbl, file.path(cfg$results_dir, paste0(cfg$platform, "_qc_summary.csv")))
readr::write_csv(qc_decision_table, file.path(cfg$results_dir, paste0(cfg$platform, "_qc_decision_table.csv")))
readr::write_csv(marker_tbl, file.path(cfg$results_dir, paste0(cfg$platform, "_cluster_markers.csv")))
readr::write_csv(annotation_map, file.path(cfg$results_dir, paste0(cfg$platform, "_annotation_review_map.csv")))
readr::write_csv(annotation_decision_table, file.path(cfg$results_dir, paste0(cfg$platform, "_annotation_decision_table.csv")))
readr::write_csv(composition_tbl, file.path(cfg$results_dir, paste0(cfg$platform, "_composition.csv")))
readr::write_csv(fov_composition_tbl, file.path(cfg$results_dir, paste0(cfg$platform, "_fov_composition.csv")))
readr::write_csv(neighbor_summary, file.path(cfg$results_dir, paste0(cfg$platform, "_neighbor_summary.csv")))
qs2::qs_save(obj, file.path(cfg$results_dir, paste0(tolower(cfg$platform), "_processed.qs2")))
```

## Output Contract

```{r}
platform_output_contract <- tibble::tribble(
  ~artifact, ~file, ~used_by, ~description,
  "processed object", paste0(tolower(cfg$platform), "_processed.qs2"), "niche, celltalk, report", "Seurat imaging object with annotations",
  "QC decision table", paste0(cfg$platform, "_qc_decision_table.csv"), "project review", "filter and segmentation decisions",
  "annotation decision table", paste0(cfg$platform, "_annotation_decision_table.csv"), "downstream labels", "marker/panel-supported annotation review",
  "FOV composition", paste0(cfg$platform, "_fov_composition.csv"), "batch review", "cell type abundance by FOV"
)

platform_output_contract
readr::write_csv(platform_output_contract, file.path(cfg$results_dir, paste0(cfg$platform, "_output_contract.csv")))
```

## Claim Evidence Table

```{r}
platform_claim_evidence <- tibble::tibble(
  claim_id = c("IMG1", "IMG2"),
  claim = c(
    "Imaging-platform annotations are marker-supported within panel limits",
    "Neighborhood patterns are robust to FOV/sample structure"
  ),
  evidence_table = c(paste0(cfg$platform, "_annotation_decision_table.csv"), paste0(cfg$platform, "_fov_composition.csv")),
  required_visual_check = c("ImageFeaturePlot marker panel", "FOV composition and neighborhood heatmap"),
  caveat = c("panel design limits biological scope", "segmentation/FOV artifacts can bias local neighborhoods"),
  status = "to_review"
)

platform_claim_evidence
readr::write_csv(platform_claim_evidence, file.path(cfg$results_dir, paste0(cfg$platform, "_claim_evidence_table.csv")))
```

## Interactive Review

- panel 是否覆盖当前组织的关键 marker？
- FOV / slide / batch 是否驱动 cluster？
- 细胞分割低质量区域是否需要剔除？
- annotation 是否有 marker 支持，还是只来自 unsupervised cluster？
- 下游进入 `spatial_niche.qmd`，不做 spot deconvolution。
