---
title: "inferCNV Pipeline"
author: "liuc"
date: "2025-01-01"
format:
  html:
    toc: true
    toc-depth: 2
---

# Overview

This notebook provides a fully-commented workflow for running **inferCNV** on an already processed Seurat object. Each chunk defaults to `eval: false`; remove or change that once you have updated the paths and parameters for your dataset.

High-level steps:

1. Load required packages and configure parallelization.
2. Load the final Seurat object (or reference an object already in memory).
3. Define malignant/reference annotations and prepare the metadata table.
4. Assemble the raw counts matrix and gene order file expected by inferCNV.
5. Construct and run the inferCNV object with optional denoising/HMM.
6. Summarize and plot CNV signals, and save results.

# 1. Setup

```{r packages}
#| label: packages
#| message: false
#| warning: false
# Step 1: load packages used in the pipeline
suppressPackageStartupMessages({
  library(Seurat)
  library(infercnv)
  library(dplyr)
  library(readr)
  library(stringr)
  library(future)
  library(tibble)
  library(rlang)
  library(ggplot2)
})
```

```{r parallel}
#| label: parallel
#| eval: false
# Step 2: optional parallelisation configuration (adjust workers to your machine)
plan("multicore", workers = 4)
```

# 2. Load Seurat Object

```{r load-object}
#| label: load-object
#| eval: false
# Step 3: load the processed Seurat object; skip if already in the environment
seurat_obj <- readRDS("data/final_seurat_object.rds")
```

```{r inspect-meta}
#| label: inspect-meta
#| eval: false
# Step 4: inspect metadata to choose annotation columns
head(seurat_obj@meta.data)
table(seurat_obj$celltype)
table(seurat_obj$sample_id)
```

# 3. Prepare Annotations

```{r define-groups}
#| label: define-groups
#| eval: false
# Step 5: specify the metadata columns and class labels
annotation_col <- "celltype"        # column that distinguishes cell states / types
malignant_groups <- c("Tumor")      # levels considered malignant
reference_groups <- c("T_cells", "B_cells", "NK_cells") # non-malignant controls

metadata_tbl <- seurat_obj@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  select(cell_id, group = !!rlang::sym(annotation_col))

stopifnot(all(malignant_groups %in% metadata_tbl$group))
stopifnot(all(reference_groups %in% metadata_tbl$group))
```

```{r write-annotation}
#| label: write-annotation
#| eval: false
# Step 6: write the annotation file (cell IDs in column 1, group labels in column 2)
annotation_path <- "inferCNV/annotations.tsv"
dir.create(dirname(annotation_path), showWarnings = FALSE, recursive = TRUE)
write_tsv(metadata_tbl, annotation_path, col_names = FALSE)
```

# 4. Prepare Expression Matrix

```{r counts-matrix}
#| label: counts-matrix
#| eval: false
# Step 7: extract the raw counts matrix (genes x cells) from the RNA assay
DefaultAssay(seurat_obj) <- "RNA"
counts_matrix <- GetAssayData(seurat_obj, slot = "counts")

# Optional: filter genes or cells prior to inferCNV (e.g., based on coverage)
# counts_matrix <- counts_matrix[rowSums(counts_matrix) > 50, ]

counts_path <- "inferCNV/counts_matrix.tsv"
write.table(counts_matrix, file = counts_path, sep = "\t",
            quote = FALSE, col.names = NA)
```

```{r gene-order}
#| label: gene-order
#| eval: false
# Step 8: build the gene order file (chromosome, gene start, gene symbol)
# Supply an existing table if you already have one; otherwise build from a GTF.
gene_order_path <- "inferCNV/gene_order.tsv"

gtf_file <- "reference/gencode.v45.basic.annotation.gtf.gz" # update to your annotation

gene_order_tbl <- read_tsv(
  gtf_file,
  comment = "#",
  col_names = c("chromosome", "source", "feature",
                "start", "end", "score", "strand", "frame", "attribute"),
  progress = FALSE
) %>%
  filter(feature == "gene") %>%
  transmute(
    gene = str_match(attribute, 'gene_name "([^"]+)"')[, 2],
    chromosome = chromosome,
    start = start
  ) %>%
  distinct(gene, .keep_all = TRUE) %>%
  arrange(chromosome, start)

# Align gene order to genes present in the counts matrix
gene_order_tbl <- gene_order_tbl %>% filter(gene %in% rownames(counts_matrix))

write_tsv(gene_order_tbl, gene_order_path, col_names = FALSE)
```

# 5. Create inferCNV Object

```{r create-infercnv}
#| label: create-infercnv
#| eval: false
# Step 9: instantiate the inferCNV object
infercnv_obj <- CreateInfercnvObject(
  raw_counts_matrix = counts_path,     # or pass counts_matrix directly
  annotations_file = annotation_path,
  delim = "\t",
  gene_order_file = gene_order_path,
  ref_group_names = reference_groups
)
```

# 6. Run inferCNV

```{r run-infercnv}
#| label: run-infercnv
#| eval: false
# Step 10: configure output directory and run the main pipeline
output_dir <- "inferCNV/results"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

infercnv_obj <- infercnv::run(
  infercnv_obj,
  cutoff = 0.1,                    # 0.1 recommended for 10x data, 1 for Smart-seq
  out_dir = output_dir,
  cluster_by_groups = TRUE,
  denoise = TRUE,
  HMM = TRUE,
  num_threads = max(1, future::nbrOfWorkers())
)
```

# 7. Post-processing

```{r load-results}
#| label: load-results
#| eval: false
# Step 11: load inferCNV outputs for downstream interpretation
cnv_matrix <- readRDS(file.path(output_dir, "infercnv.observations.rds"))
reference_matrix <- readRDS(file.path(output_dir, "infercnv.references.rds"))

# Example: compute average CNV signal per group
cnv_df <- as.data.frame(cnv_matrix)
cnv_df$gene <- rownames(cnv_matrix)
```

```{r quick-plot}
#| label: quick-plot
#| eval: false
# Step 12: quick visualization of chromosome-level CNV scores
infercnv::plot_cnv(
  infercnv_obj,
  out_dir = file.path(output_dir, "plots"),
  plot_per_chr = TRUE,
  title = "inferCNV CNV heatmap"
)
```

# 8. Save Updated Object

```{r save-objects}
#| label: save-objects
#| eval: false
# Step 13: optionally attach inferCNV results back to the Seurat object
seurat_obj@misc$infercnv <- list(
  cnv_matrix = cnv_matrix,
  reference_matrix = reference_matrix,
  params = list(
    malignant = malignant_groups,
    reference = reference_groups
  )
)

saveRDS(seurat_obj, file = "inferCNV/seurat_with_cnv.rds")
```

# Notes

- Ensure the cell names in the Seurat object and annotation file match exactly.
- If you prefer to avoid intermediate files, supply matrices and data frames directly to `CreateInfercnvObject`; see `?CreateInfercnvObject` for accepted input types.
- Review the inferCNV documentation for advanced settings such as Bayesian mixture models, multi-sample comparisons, and alternative plotting modes.
